---
title: 'Reproducible Research: Peer Assessment 1'
author: "Abhinav Khandelwal"
date: "2024-02-29"
output: html_document
---

## Loading and preprocessing the data

To load the data in R, download the file from [Activity_Data](https://github.com/AK282331/RepData_PeerAssessment1/blob/master/activity.zip) and then save the file in that directory which you have set as a current working directory in R.

```{r Loading the data, echo=TRUE}
activity = read.csv(file = "C:\\Users\\Abhinav Khandelwal\\Desktop\\R_project\\R_Learning_Coursera\\activity.csv")
```
```{r Preprocessing the data, echo = TRUE}

### Converting date from character to date format

activity$date = as.Date(activity$date,format = "%Y-%m-%d")
head(activity)
```

```{r,installing dplyr package,echo=FALSE,results='hide',attr.warning=NA}
# Set the CRAN mirror
options(repos = "https://cloud.r-project.org/")
## install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

## What is mean total number of steps taken per day?

**1. Calculate the total number of steps taken per day.**

```{r}
activity = activity[complete.cases(activity),]
total_steps = activity %>% group_by(date) %>% summarise(sum(steps,na.rm = T)) %>%  as.data.frame()
total_steps
```

**2. Make a histogram of the total number of steps taken each day.**

```{r histogram of total number of steps each day}

hist(total_steps$`sum(steps, na.rm = T)`,xlab = "Total Steps Per Day",main = "Histogram of total number of steps\ntaken each day")

```

**3.Calculate and report the mean and median of the total number of steps taken per day.**

```{r Calculating Mean and Median}
total_steps_mean = mean(total_steps$`sum(steps, na.rm = T)`)
total_steps_median = median(total_steps$`sum(steps, na.rm = T)`)
cat ("total_steps_mean = ",total_steps_mean)
cat ("total_steps_median = ",total_steps_median)
```

## What is the average daily activity pattern?

**1.Make a time series plot (i.e.type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

For the above question we need to first calculate the averages of steps taken on the basis of intervals.

```{r Calculating average of intervals}
interval_avg = tapply(activity$steps,factor(activity$interval),mean)
interval_avg = data.frame(intervals = unique(activity$interval),average = interval_avg)
head(interval_avg)
```


``` {r Plotting Time-Series of 5-minute interval and average number of steps taken }
plot(interval_avg$intervals,interval_avg$average, type = "l",main = "Time-Series Plot of \n5-minute interval and average number of steps taken",xlab = "5-Minute Intervals",ylab = "Average Number of Steps")
```

**2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r Maximun 5-Minute Interval}
interval_avg[which.max(interval_avg$average),]

```

## Imputing missing values

**1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

``` {r}
activity2 = read.csv(file = "C:\\Users\\Abhinav Khandelwal\\Desktop\\R_project\\R_Learning_Coursera\\activity.csv")

head(activity2)
```

```{r total number of rows with NAs}
Nas = sum(!complete.cases(activity2))
Nas
```

**2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

Here my strategy is to use the average of 5-minute interval in place of Na's.

```{r Replace NAs with mean of 5-minute interval}

activity_null = filter(activity2,is.na(steps))
for (i in unique(activity_null$interval)){
    avg = interval_avg[interval_avg$intervals==i,2]
    activity_null$steps= replace(activity_null$steps,activity_null$interval == i,avg)
    
}
head(activity_null)

```

**3.Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r New Dataset with Missing Data filled}

for (i in unique(activity2$interval)){
        val = interval_avg[interval_avg$intervals==i,2]
        activity2$steps = replace(activity2$steps,activity2$interval==i & is.na(activity2$steps),val)  }
   
head(activity2)                                     
```

**4.Make a histogram of the total number of steps taken each day**

```{r histogram of the total number of steps taken each day}
activity2$date = as.Date(activity2$date,format = "%Y-%m-%d")
total_steps_new = activity2 %>% group_by(date) %>% summarise(sum(steps))
hist(total_steps_new$`sum(steps)`,xlab = "Total Steps Per Day",main = "Histogram of total number of steps\ntaken each day")

```

**4. and Calculate and report the mean and median total number of steps taken per day.**  
```{r Calculating Mean and Median on new data}
total_steps_new_mean = mean(total_steps_new$`sum(steps)`)
total_steps_new_median = median(total_steps_new$`sum(steps)`)

cat("total_steps_new_mean = ",total_steps_new_mean)
cat("total_steps_new_median = ",total_steps_new_median)
```


**4.Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**

When we had the data with NAs then the mean was`r total_steps_mean` and median was `r total_steps_median`. Now after imputing the NA's we have mean `r total_steps_new_mean` and median `r total_steps_new_median`. With this we can see that there is no change in the mean but there is a change in the median of 1.19 steps and npw we have normal distribution.

## Are there differences in activity patterns between weekdays and weekends?

**1.Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.**

```{r creating new factor variable with weekday and weekend levels}
activity2 = mutate(activity2,days = weekdays(date),weekday_end = case_when(days %in% c("Monday","Tuesday","Wednesday","Thursday","Friday")~"weekday",days %in% c("Saturday", "Sunday") ~ "weekend")) %>% select(-4) %>% as.data.frame()

activity2$weekday_end = factor(activity2$weekday_end)

head(activity2)
```

**2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.**

``` {r Plotting Time Series plot}
## Creating a new data frame which contains the interval and their averages

interval_avg_new = activity2 %>% group_by(interval,weekday_end) %>% summarise(mean(steps)) %>% as.data.frame() %>% rename(average_steps = 'mean(steps)')

## Creating Time-Series Plot

gg = ggplot(interval_avg_new,aes(interval,average_steps,col = weekday_end))
gg+geom_line() + facet_grid(weekday_end~.) + ggtitle(label = "Time-Series Plot of the 5-minute interval and the average number of steps taken \nacross all weekday days or weekend days") + xlab("5-Minute Intervals") + ylab("Average Steps Taken")

            




